% {\footnotesize\em This page is optional.
% It's in the table of contents and it's labeled 'ACKNOWLEDGMENTS' even
% though the spelling `ACKNOWLEDGEMENTS' is also correct.
% This page should be at the end of the preface, if one exists, or a
% separate page, if no preface is used.}

% Recognize faculty, friends, family but not the typist, editor or persons
% paid to produce the manuscript from draft copy. Recognize funding,
% special permission to reproduce figures and copyrighted material.

% A special thank you to ...

% And finally, my wonderful wife, without whose innumerable sacrfices and enduring support I could not have succeeded.

A big thank you to my advisor, Matthew Flatt, for being willing to take on a clueless Masters student. Matthew sees things so clearly. I am deeply appreciative of his guidance and perspective. Thank you to the other members of my committee: Vivek Srikumar and Jonathan Cohen. Their input brought an incredibly diverse set of possibilities to my attention; I regret not having enough time to develop all of them more fully.

Thanks to the other students involved as well. Thanks to Annie Cherkaev for always being willing to answer my questions and taking the time to help me understand. Thanks to Sebastian Musslick for being the patient guinea pig and providing valuable feedback every step of the way.

A heartfelt thank you to my friends and teammates at FamilySearch: Jim, Chris, Bob, Joseph, Barrett, Brent, and many others. They all pulled extra weight to support me when the academic load was particularly heavy, and I credit them for this achievement as much as anyone else. I couldn't ask for a better team.

Thanks to Mom and Dad for always lifting me up, and never letting me believe that I wasn't smart enough.

Lastly, and lest there be any doubt, most importantly, thank you to my extraordinary wife, Brittney. Over the last three years, she has played the single mother to our children more times than I can count while I stayed late at work or the library. As long as I've known her, she has always been willing to sacrifice her own interests to benefit her friends and family. I certainly could not have succeeded were she absent. I love her dearly.

